---
output: github_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",  # <— wichtig, damit Bilder gespeichert werden
  fig.width = 7,
  fig.height = 6,
  dpi = 150
)
library(almmr)
library(terra)
library(sf)
```

# almmr – Archaeological Landscape Movement Modeling in R

The **almmr** package provides tools for modeling movement, accessibility, and territorial structure in archaeological landscapes.  
It builds on digital elevation models (DEMs) and implements cost-surface modeling, least-cost path analysis (TPLA), and spatial boundary reconstruction (SBR) methods.  
All examples below use a 25 m DEM of the Heuneburg region in Baden-Württemberg, Germany, included with the package.

---

## Citation

Please cite **almmr** if you use it in your research; run `citation("almmr")`
in R or see `CITATION.cff` for the reference. The underlying dissertation may
additionally be cited for the methodological background.
---

## Installation

**almmr** is installed from GitHub with the **devtools** package, so install
that first if you don't have it:

```r
# devtools is required to install from GitHub
install.packages("devtools")

# Install almmr from GitHub
devtools::install_github("j-abele/almmr")

library(almmr)
library(terra)
library(sf)
```

---

## Cost Surface Generation

Most analyses build a cost surface first with create_cost_surface() and pass the resulting almmr_cs object on; this is required for perform_tpla(), perform_tpla_line_based() and sbr_network(). Others build it internally from the DEM — lcsc_territory(), for instance, only needs the DEM — so calling create_cost_surface() explicitly is not always necessary.

```{r cost-surface, message=FALSE, warning=FALSE, fig.alt="DEM of the Heuneburg region shown over a hillshade"}
# Load DEM and create hillshade
r  <- almmr::load_dem()
hs <- almmr::create_hillshade(r)

# Create cost surface using Tobler's Hiking Function.
# The result is an 'almmr_cs' object carrying the DEM, edge weights and
# parameters; pass it to perform_tpla(), lcsc_territory() and sbr_network().
cs <- almmr::create_cost_surface(
  dem          = r,
  costFunction = "ToblersHikingFunction"
)

# Plot the DEM underlying the cost surface on the hillshade
plot(hs, col = gray.colors(256, 0.1, 1), legend = FALSE, main = "DEM / Cost Surface (Tobler)")
plot(cs$dem, add = TRUE, alpha = 0.5)
```

---

## TPLA (Total Passability Landscape Analysis)

TPLA stands for Total Passability Landscape Analysis. With the help of least-cost paths and a density calculation, regions with high route or movement potential are calculated within a circular analysis area.


```{r tpla, message=FALSE, warning=FALSE, fig.alt="TPLA movement-potential density around the Heuneburg"}
# Heuneburg as center point (in the DEM CRS)
heuneburg <- terra::vect(cbind(530657, 5326988), crs = terra::crs(r))

# Compute Total Passability Landscape Analysis for the Heuneburg micro-region.
# perform_tpla() reprojects the center point to the cost-surface CRS if needed.
tpla_heune <- almmr::perform_tpla(
  cost_surface       = cs,
  center_point       = heuneburg,
  radius_tpla        = 4000,
  number_of_points   = 30,
  sigma_density_calc = 90,
  keep_lines         = FALSE
)

# Plot density on hillshade
plot(hs, col = gray.colors(256, 0.1, 1), legend = FALSE, main = "Total Passability Landscape Analysis")
# Mask out cells with near-zero values (remove low-density / non-passable areas)
tpla_heune_mask <- terra::mask(tpla_heune, tpla_heune > 0.04, maskvalues = FALSE)
plot(tpla_heune_mask, add = TRUE, alpha = 0.7)
plot(heuneburg, col = "red", pch = 16, add = TRUE)

```

---

## LCSC Territory - Potential settlement area

The function generates theoretical territories or catchment areas based on a cost surface for one or more locations. The catchment areas are calculated on the basis of a certain period of time, which, starting from a central location, the movement can take place in all directions depending on the costs.


```{r lcsc, message=FALSE, warning=FALSE, fig.alt="LCSC catchment territories around ten random sites"}
# Create 10 random settlement points
set.seed(7)
points_sf <- sf::st_sample(sf::st_as_sf(terra::as.polygons(r)), 10)
points_sf <-  terra::vect(points_sf)


# Run LCSC territory calculation (returns a SpatVector of polygons)
territories <- almmr::lcsc_territory(
  dem           = r,
  sites         = points_sf,
  movement_time = 5,
  max_speed     = 6
)

# Plot territories on hillshade
plot(hs, col = gray.colors(256, 0.1, 1), legend = FALSE, main = "LCSC Territories (n = 10)")
plot(territories, col = hcl.colors(10, "Spectral", alpha = 0.5), add = TRUE)
plot(points_sf, col = "black", pch = 16, add = TRUE)
```

---

## SBR_network - Site-based (theoretical) route network (sbr_network)

The `sbr_network()` function constructs a theoretical route network between given site points based on the underlying cost surface. The selection of the next neighboring site is determined iteratively using accumulated cost surfaces, connecting locations via the least-cost paths and thus simulating potential communication or movement corridors within a landscape.

⚠️ **Note:** This is a computationally intensive process.  
Even with the small example DEM included in this package, depending on your hardware and the sampling density (`steps_points`), it may take several minutes to complete.  
So it’s a good idea to grab a coffee and let R do the heavy lifting ☕.


```{r sbr, message=FALSE, warning=FALSE, fig.alt="Site-based route network connecting sites via least-cost paths"}
# Create 10 random points in the DEM extent
set.seed(7)
network_pts <- sf::st_sample(sf::st_as_sf(terra::as.polygons(r)), 10)
network_pts <- terra::vect(network_pts)

# Plot the points; two of them serve as endpoints for a reference route
plot(hs, col = gray.colors(256, 0.1, 1), legend = FALSE, main = "SBR Network and Reference Path")
plot(network_pts, col = "black", pch = 16, add = TRUE)
plot(network_pts[c(2, 4)], col = "green", pch = 16, add = TRUE)

# Reference least-cost path between the two endpoints
# (replaces the former gdistance::shortestPath(); returns an sf LINESTRING)
ref_line <- almmr::compute_lcp(
  dem         = r,
  origin      = network_pts[2],
  destination = network_pts[4],
  cs          = cs
)
plot(sf::st_geometry(ref_line), col = "black", add = TRUE)

# Build the site-based route network from the remaining points.
# sbr_network() requires an eager cost surface (create_cost_surface(lazy = FALSE)).
sbr_net <- almmr::sbr_network(
  sites        = network_pts[-c(2, 4)],
  lines        = ref_line,
  cost_surface = cs,
  steps_points = 150
)

# Plot the resulting network
plot(sf::st_geometry(sbr_net), col = "darkgreen", add = TRUE)


```



---

## Provenance and development

The methodological approaches for Total Passability Landscape Analysis (TPLA),
least-cost settlement catchment (LCSC) territories, and site-based route
network (SBR) construction, together with their implementation as R functions,
are the author's own work, developed by Jonas Abele as part of the doctoral
dissertation *Der „Fürstensitz“ Heuneburg in seinem regionalen Kontext. Eine
landschaftsarchäologische Untersuchung zur Entwicklung hierarchischer
Siedlungsstrukturen* (University of Tübingen, 2024) and in part inspired by
existing literature. The methods and results are described in
detail in the dissertation; the original implementation was built on the
[`gdistance`](https://cran.r-project.org/package=gdistance) package.

For publishing this code after completion of the doctorate — in particular for
error checking, the migration from `gdistance` to a native `terra`/`igraph`
implementation, and the creation of parts of the documentation — AI
language-model (LLM) assistance was used.

---

## License

Code licensed under the GNU General Public License v3.0 (GPL-3).  
Example data (Heuneburg DEM) © LGL Baden-Württemberg
